Four-dimensional emittance measurement at the Spallation Neutron Source

### 2022-07-24 main.tex
Rename figures

A comprehensive study of errors in the multi-optics 4D emittance measurement was completed at the SwissFEL Injector Test Facility (SITF) by Prat and Aiba in \cite{Prat2014}. They considered errors in the measured moments, quadrupole field and alignment errors, beam energy errors, beam mismatch at the reconstruction point, and dispersion/chromaticity \cite{Mostacci2012}, concluding that the method remained accurate and reporting $< 5\%$ uncertainty in the intrinsic emittances. We initially performed similar studies using envelope tracking to estimate the reconstruction errors in the RTBT \cite{Hoover2021-IPAC}. We considered uncertainty in the beam moments ($\pm 5\%$) quadrupole tilt angles ($\pm\%$ 1 mrad), quadrupole field strengths ($\pm$ 1\%), beam kinetic energy ($\pm$ 3 MeV for a 1 GeV beam), and beam Twiss parameters at the reconstruction point ($\pm$ 5\%), with all other effects turned off. The emittances were successfully reconstructed with systematic and random errors of a few percent. Most of these uncertainties had only a small effect on the reconstructed intrinsic emittances, with the largest coming from beam mismatch and errors in the measured beam size. Therefore, we only consider these sources of uncertainty going forward.\footnote{Space charge forces, which can render the method invalid for high-perveance beams \cite{Anderson2002}, can be neglected since the space charge tune shift in the ring is around 3\% and the distance from the reconstruction location to the measurement location is much less than the length of the ring.} From repeated wire-scanner measurements with the same machine setup (see Fig.~\ref{fig:3}), we estimated the error in the measured moments to be only 3\%. This 3\% noise will be assumed throughout the rest of the analysis.
%
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{profiles.pdf}
    \caption{Measured profiles of a production beam at WS20 in the RTBT.}
    \label{fig:3}
\end{figure}
%

@@

To investigate the sensitivity of the fixed-optics method, we generated a covariance matrix with $\varepsilon_1 = \varepsilon_x$ = 32 mm~mrad and $\varepsilon_2 = \varepsilon_y$ = 25 mm~mrad, close to the measured values, and with Twiss parameters matched to the lattice. This covariance matrix was tracked to the wire-scanners using the transfer matrices from the fifth step in the scan of Fig.~\ref{fig:2}, and the reconstruction was performed many times with 3\% random noise added to the $\langle{xx}\rangle$, $\langle{yy}\rangle$, and $\langle{uu}\rangle$ moments. (We ignored any uncertainty in the transfer matrix elements, which were expected to be small.) This resulted in a large fraction of failed trials, but some successful trials. Fig.~\ref{fig:4} shows the emittances in the successful trials.
%
\begin{figure}[!b]
    \centering
    \includegraphics[width=\columnwidth]{fig3.pdf}
    \caption{Simulated emittance measurement using the nominal optics in the RTBT.}
    \label{fig:4}
\end{figure}
%
Unlike the apparent emittances, the intrinsic emittances are strongly correlated and are not centered on the correct values. Our primary concern is to reduce the latter effect, which we refer to as bias.

@@

We performed a similar modification of the RTBT optics, using these previous studies as a guide. Recall the two available knobs: the two power supplies in the wire-scanner region used to control the phase advances at the final wire-scanner. To search for a better set of optics, we varied these phase advances ($\mu_x$, $\mu_y$) in a $90\degree$ window around their nominal values ($\mu_{x0}$, $\mu_{y0}$). At each setting, a matched covariance matrix with $\varepsilon_1 = \varepsilon_2 = \varepsilon_x = \varepsilon_y$ = 20 mm~mrad was generated and tracked to the wire-scanner locations. The reconstruction was again simulated with 3\% random noise added to the ``measured" moments, repeating over a few thousand trials. The mean and standard deviation of the emittances were calculated over the successful trials. The difference between the mean emittances and the true emittances, which we refer to as the \textit{bias}, is plotted for each set of optics in the top row of Fig.~\ref{fig:5}, as well as the standard deviations in the bottom row.
%
\begin{figure}[]
    \centering
    \includegraphics[width=\columnwidth]{fig4.pdf}
    \caption{Simulated bias and standard deviation of the intrinsic emittances in the RTBT as a function of the phase advances at WS24.}
    \label{fig:5}
\end{figure}
%
%
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[b]{0.92\textwidth}
       \includegraphics[width=\linewidth]{fig5.pdf}
       \caption{}
       \label{fig:6a} 
    \end{subfigure}
    \begin{subfigure}[b]{0.92\textwidth}
       \includegraphics[width=\linewidth]{fig5b.pdf}
       \caption{}
       \label{fig:6b}
    \end{subfigure}
    \caption{(a) Bias and standard deviations of $\varepsilon_1$ (pink) and $\varepsilon_2$ (blue) from simulated reconstructions in the RTBT as the beam Twiss parameters $\left\{\alpha_x, \alpha_y, \beta_x, \beta_y\right\}$ are scanned within a four-dimensional grid around their nominal values (as defined by the lattice). Each point in each subfigure corresponds to a simulation with different Twiss parameters. The beam emittances are fixed in each column, varying from no cross-plane correlation on the far left to large cross-plane correlation on the far right.  (b) For each subfigure in (a), from left to right, we plot the distribution of intrinsic emittances generated by the random trials for one set of Twiss parameters (greyscale). Each blue shaded band is centered at the mean of the distribution one axis, with a width equal to twice the standard deviation along that axis. The red lines show the true values of the emittances.}
    \label{fig:6}
\end{figure*}
%
Settings that produced no successful trials appear as white cells. The apparent emittances are not displayed because they remained within 1\% of their true values at every optics setting. Modifying the optics so that $\mu_x - \mu_{x0} = 45\degree$ and $\mu_y - \mu_{y0} = -45\degree$ reduces the bias to $\approx 7\%$ and the standard deviation to $\approx 5\%$. The fraction of failed fits, which is very large along the diagonal in the figure, is reduced to zero at this setting.

@@

We first display the measured emittance evolution of a 1 GeV beam produced by the standard SNS painting method.\footnote{The standard painting method is often referred to as "correlated painting". The distance between the injected and circulating beam is varied as the square root of time while the angle between the beams is fixed at zero. Note that ``correlated painting" tends not to produce any cross-plane correlation in the beam. See \cite{Wang1999, Hotchi2020}.} The beam was accumulated over 500 injected turns; it was extracted and measured every 50 turns during accumulation. The reconstructed emittances throughout accumulation are shown in Fig.~\ref{fig:7}a.
%
\begin{figure}[t!]
    \centering
    \includegraphics[width=\columnwidth]{fig6.pdf}
    \caption{Measured emittances vs. number of injected turns in the SNS ring. (a) correlated (production) painting at 1 GeV; (b) attempted elliptical painting at 1 GeV; (c)-(d) elliptical painting at 0.8 GeV. In (d), the painted beam size was increased and the beam intensity was decreased relative to (c).}
    \label{fig:7}
\end{figure}
%
As expected, the beam did not develop any significant cross-plane correlation. The beam Twiss parameters were essentially matched at the nominal optics at the RTBT entrance, so bias in the measurement was not a concern.

@@

%
\begin{figure*}[!t]
    \centering
    \begin{subfigure}{0.48\textwidth}
        \includegraphics[width=\linewidth]{fig10a.pdf}
        \caption{}
        \label{fig:10a} 
    \end{subfigure}
    \hspace{0.4cm}
    \begin{subfigure}{0.48\textwidth}
        \includegraphics[width=\linewidth]{fig10b.pdf}
        \caption{}
        \label{fig:10b}
    \end{subfigure}
    \caption{Emittance distribution used to produce the error bars in Fig.~\ref{fig:7}d at (a) 150 turns (b) 400 turns. The points were generated by reconstructing the covariance matrix from perturbations of the measured moments. The measured values are plotted in red.}
    \label{fig:10}
\end{figure*}
%
%
\begin{figure}[t!]
    \centering
    \includegraphics[width=\columnwidth]{fig9.pdf}    \caption{2D projections of the reconstructed covariance ellipsoid as a function of turn number during injection in the SNS ring. The measured emittances are shown in Fig.~\ref{fig:7}d.}
    \label{fig:9}
\end{figure}
%
In this appendix, we offer two clarifications on the measurements in Fig.~\ref{fig:7}d. First, Fig.~\ref{fig:10} displays the distribution of apparent and intrinsic emittances used to generate the error bars in Fig.~\ref{fig:7}d at turn 150 and turn 400. Again, these emittances were obtained by repeating the reconstruction as 3\% noise was added to the measured $\langle{xx}\rangle$, $\langle{yy}\rangle$ and $\langle{xy}\rangle$. Notice that in (a), when the cross-plane correlation is small, the mean of the distribution is not exactly equal to the measured emittances, leading to the asymmetric error bars in Fig.~\ref{fig:7}d.

Second, since the intrinsic emittances are complicated functions of the second-order beam moments, it is not always obvious how changes in the intrinsic emittances affect the beam distribution. Therefore, in Fig.~\ref{fig:9}, we have plotted the 2D projections of the covariance ellipsoid for each measurement in Fig.~\ref{fig:7}d. The change in intrinsic emittances manifests in a tilting of the ellipses in planes other than $x$-$x'$ and $y$-$y'$.

>>>>

A comprehensive study of errors in the multi-optics 4D emittance measurement was completed at the SwissFEL Injector Test Facility (SITF) by Prat and Aiba in \cite{Prat2014}. They considered errors in the measured moments, quadrupole field and alignment errors, beam energy errors, beam mismatch at the reconstruction point, and dispersion/chromaticity \cite{Mostacci2012}, concluding that the method remained accurate and reporting $< 5\%$ uncertainty in the intrinsic emittances. We initially performed similar studies using envelope tracking to estimate the reconstruction errors in the RTBT \cite{Hoover2021-IPAC}. We considered uncertainty in the beam moments ($\pm 5\%$) quadrupole tilt angles ($\pm\%$ 1 mrad), quadrupole field strengths ($\pm$ 1\%), beam kinetic energy ($\pm$ 3 MeV for a 1 GeV beam), and beam Twiss parameters at the reconstruction point ($\pm$ 5\%), with all other effects turned off. The emittances were successfully reconstructed with systematic and random errors of a few percent. Most of these uncertainties had only a small effect on the reconstructed intrinsic emittances, with the largest coming from beam mismatch and errors in the measured beam size. Therefore, we only consider these sources of uncertainty going forward.\footnote{Space charge forces, which can render the method invalid for high-perveance beams \cite{Anderson2002}, can be neglected since the space charge tune shift in the ring is around 3\% and the distance from the reconstruction location to the measurement location is much less than the length of the ring.} From repeated wire-scanner measurements with the same machine setup (see Fig.~\ref{fig:3}), we estimated the error in the measured moments to be only 3\%. This 3\% noise will be assumed throughout the rest of the analysis.
%
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{fig3.pdf}
    \caption{Measured profiles of a production beam at WS20 in the RTBT.}
    \label{fig:3}
\end{figure}
%

@@

To investigate the sensitivity of the fixed-optics method, we generated a covariance matrix with $\varepsilon_1 = \varepsilon_x$ = 32 mm~mrad and $\varepsilon_2 = \varepsilon_y$ = 25 mm~mrad, close to the measured values, and with Twiss parameters matched to the lattice. This covariance matrix was tracked to the wire-scanners using the transfer matrices from the fifth step in the scan of Fig.~\ref{fig:2}, and the reconstruction was performed many times with 3\% random noise added to the $\langle{xx}\rangle$, $\langle{yy}\rangle$, and $\langle{uu}\rangle$ moments. (We ignored any uncertainty in the transfer matrix elements, which were expected to be small.) This resulted in a large fraction of failed trials, but some successful trials. Fig.~\ref{fig:4} shows the emittances in the successful trials.
%
\begin{figure}[!b]
    \centering
    \includegraphics[width=\columnwidth]{fig4.pdf}
    \caption{Simulated emittance measurement using the nominal optics in the RTBT.}
    \label{fig:4}
\end{figure}
%
Unlike the apparent emittances, the intrinsic emittances are strongly correlated and are not centered on the correct values. Our primary concern is to reduce the latter effect, which we refer to as bias.

@@

We performed a similar modification of the RTBT optics, using these previous studies as a guide. Recall the two available knobs: the two power supplies in the wire-scanner region used to control the phase advances at the final wire-scanner. To search for a better set of optics, we varied these phase advances ($\mu_x$, $\mu_y$) in a $90\degree$ window around their nominal values ($\mu_{x0}$, $\mu_{y0}$). At each setting, a matched covariance matrix with $\varepsilon_1 = \varepsilon_2 = \varepsilon_x = \varepsilon_y$ = 20 mm~mrad was generated and tracked to the wire-scanner locations. The reconstruction was again simulated with 3\% random noise added to the ``measured" moments, repeating over a few thousand trials. The mean and standard deviation of the emittances were calculated over the successful trials. The difference between the mean emittances and the true emittances, which we refer to as the \textit{bias}, is plotted for each set of optics in the top row of Fig.~\ref{fig:5}, as well as the standard deviations in the bottom row.
%
\begin{figure}[]
    \centering
    \includegraphics[width=\columnwidth]{fig5.pdf}
    \caption{Simulated bias and standard deviation of the intrinsic emittances in the RTBT as a function of the phase advances at WS24.}
    \label{fig:5}
\end{figure}
%
%
\begin{figure*}[t!]
    \centering
    \begin{subfigure}[b]{0.92\textwidth}
       \includegraphics[width=\linewidth]{fig6a.pdf}
       \caption{}
       \label{fig:6a} 
    \end{subfigure}
    \begin{subfigure}[b]{0.92\textwidth}
       \includegraphics[width=\linewidth]{fig6b.pdf}
       \caption{}
       \label{fig:6b}
    \end{subfigure}
    \caption{(a) Bias and standard deviations of $\varepsilon_1$ (pink) and $\varepsilon_2$ (blue) from simulated reconstructions in the RTBT as the beam Twiss parameters $\left\{\alpha_x, \alpha_y, \beta_x, \beta_y\right\}$ are scanned within a four-dimensional grid around their nominal values (as defined by the lattice). Each point in each subfigure corresponds to a simulation with different Twiss parameters. The beam emittances are fixed in each column, varying from no cross-plane correlation on the far left to large cross-plane correlation on the far right.  (b) For each subfigure in (a), from left to right, we plot the distribution of intrinsic emittances generated by the random trials for one set of Twiss parameters (greyscale). Each blue shaded band is centered at the mean of the distribution one axis, with a width equal to twice the standard deviation along that axis. The red lines show the true values of the emittances.}
    \label{fig:6}
\end{figure*}
%
Settings that produced no successful trials appear as white cells. The apparent emittances are not displayed because they remained within 1\% of their true values at every optics setting. Modifying the optics so that $\mu_x - \mu_{x0} = 45\degree$ and $\mu_y - \mu_{y0} = -45\degree$ reduces the bias to $\approx 7\%$ and the standard deviation to $\approx 5\%$. The fraction of failed fits, which is very large along the diagonal in the figure, is reduced to zero at this setting.

@@

We first display the measured emittance evolution of a 1 GeV beam produced by the standard SNS painting method.\footnote{The standard painting method is often referred to as "correlated painting". The distance between the injected and circulating beam is varied as the square root of time while the angle between the beams is fixed at zero. Note that ``correlated painting" tends not to produce any cross-plane correlation in the beam. See \cite{Wang1999, Hotchi2020}.} The beam was accumulated over 500 injected turns; it was extracted and measured every 50 turns during accumulation. The reconstructed emittances throughout accumulation are shown in Fig.~\ref{fig:7}a.
%
\begin{figure}[t!]
    \centering
    \includegraphics[width=\columnwidth]{fig7.pdf}
    \caption{Measured emittances vs. number of injected turns in the SNS ring. (a) correlated (production) painting at 1 GeV; (b) attempted elliptical painting at 1 GeV; (c)-(d) elliptical painting at 0.8 GeV. In (d), the painted beam size was increased and the beam intensity was decreased relative to (c).}
    \label{fig:7}
\end{figure}
%
As expected, the beam did not develop any significant cross-plane correlation. The beam Twiss parameters were essentially matched at the nominal optics at the RTBT entrance, so bias in the measurement was not a concern.

@@

%
\begin{figure*}[!t]
    \centering
    \begin{subfigure}{0.48\textwidth}
        \includegraphics[width=\linewidth]{fig9a.pdf}
        \caption{}
        \label{fig:9a} 
    \end{subfigure}
    \hspace{0.4cm}
    \begin{subfigure}{0.48\textwidth}
        \includegraphics[width=\linewidth]{fig9b.pdf}
        \caption{}
        \label{fig:9b}
    \end{subfigure}
    \caption{Emittance distribution used to produce the error bars in Fig.~\ref{fig:7}d at (a) 150 turns (b) 400 turns. The points were generated by reconstructing the covariance matrix from perturbations of the measured moments. The measured values are plotted in red.}
    \label{fig:9}
\end{figure*}
%
%
\begin{figure}[b!]
    \centering
    \includegraphics[width=\columnwidth]{fig10.pdf}    
    \caption{2D projections of the reconstructed covariance ellipsoid as a function of turn number during injection in the SNS ring. The measured emittances are shown in Fig.~\ref{fig:7}d.}
    \label{fig:10}
\end{figure}
%
In this appendix, we offer two clarifications on the measurements in Fig.~\ref{fig:7}d. First, Fig.~\ref{fig:9} displays the distribution of apparent and intrinsic emittances used to generate the error bars in Fig.~\ref{fig:7}d at turn 150 and turn 400. Again, these emittances were obtained by repeating the reconstruction as 3\% noise was added to the measured $\langle{xx}\rangle$, $\langle{yy}\rangle$ and $\langle{xy}\rangle$. Notice that in (a), when the cross-plane correlation is small, the mean of the distribution is not exactly equal to the measured emittances, leading to the asymmetric error bars in Fig.~\ref{fig:7}d.

Second, since the intrinsic emittances are complicated functions of the second-order beam moments, it is not always obvious how changes in the intrinsic emittances affect the beam distribution. Therefore, in Fig.~\ref{fig:10}, we have plotted the 2D projections of the covariance ellipsoid for each measurement in Fig.~\ref{fig:7}d. The change in intrinsic emittances manifests in a tilting of the ellipses in planes other than $x$-$x'$ and $y$-$y'$.
